---
title: Why are NYC apartment prices not falling more?
author: Seth Ariel Green
date: '2020-05-21'
slug: why-are-nyc-apartment-prices-not-falling-more
---

(This is a test blogpost so I'm going to keep it short and sweet)

### Why haven't prices fallen?

I have an apartment in New York. I don't currently reside there and will move out when the lease ends (in about a month). It seems that Manhattan is somewhere you wouldn't want to be right now if you could be anywhere else. 

So why aren't real estate prices cheaper? 

Browsing streeteasy, I get the sense that prices are reduced _slightly_ from what they were three months ago, despite it being the epicenter of a pandemic, and therefore:

* the things that make New York special (food, opera, whatever your thing is) are basically completely in abeyance;
* everyday things like exiting and entering your building are much more fraught than they used to be;
* if people are the ultimate amenity, a lot of them are gone.

So, if I didn't know better, I'd expect there to be a mass exodus of renters, and for prices to reflect the suddenly high supply-to-demand ratio. What gives? 

Possibly, prices might fall a lot in the summer (assuming lots of people, like me, end their leases and landlords find it difficult to replace them).

Another possibility: rent stabilization laws put a floor on how much many landlords are willing to let prices drop.

### Working through a simple example:

* About a million apartments in New York are rent-stabilized if not rent controlled. 
* Let's say you (the landlord) have a rent-stabilized apartment that, in normal times, rents for $2000/month.
* The actual market-clearing price for the apartment during these weird times is $1000/month.
* You estimate that, over the next 10 years, you'll only be able to raise the rent an average of 5% per year (you can't get a perfect estimate of this because legislation is liable to change the rate).
* If an apartment is vacant, you can raise rents by up to 20%.
* In 2 years time, you expect the apartment to be worth $2000/month again.

### What would you do? 

On the one hand, you could leave the apartment up on Streeteasy for \$2000/month, even though you think that it will stay vacant for two years. On paper, you're going to lose \$48,000 in that time -- but remember, the true, market-clearing price for the aparmtent is $1000/month, so you're actually losing \$24,000. In two years time, you expect to be able to rent it for \$2000/month and increase the rent by 5% every year for the following 8 years.

On the other, you could reduce the price to $1000 and then aim to raise the price by 5% every year for the next ten years. 

In the first scenario, your income for the 10 year period is (this is going to fast and loose, I'm sure there's some `lapply` way to do it better):
```{r}
rent_calc <- function(base_amount, increase_rate, years) {
  rent <- (base_amount * 12)
  appreciation <- 1 + increase_rate
  for(i in 1:(years-1)){
    rent <- rent + (base_amount * appreciation^i * 12)
  }
  print(rent)
}

rent_calc(2000, 0.05, 8)
```
vs, in the second scenario:

```{r}
rent_calc(1000, 0.05, 10)
```


\$150,934.7 -- it's not even close. You lose about \$78,000 over the course of 10 years if you put the apartment on the market now. 

What if the maximum appreciation rate were 10%, and in 10 years time, you thought that the apartment would still be worth $2000? 

Putting it on the market for $1000 now, you'd get:
```{r}
rent_calc(1000, .1, 10)

```

Leaving it vacant and then charging $2000/month every year for the following 8:
```{r}
2000 * 12 * 8
```

You're still slightly better off leaving it off the market.

Am I right? I played a little fast and loose with the details here and I am not totally sure I got rent stabilization right :) But let's say I'm right; if so:

### A prediction and a suggestion

1) A lot of apartments will come on the market this year with provisions like "3 months free" (effectively reducing the monthy rent by a lot, but, as far as I know, a provision not covered by typical rent stabilization provisions)

2) If you like the looks of an apartment, and you have a good relationship with the landlord (or think you could develop one), you could strike a sweet deal here. For this hypothetical \$2000 apartment, tell them that you're willing to pay \$1000 a month -- one suspects that lanldords are feeling a little cash-strapped at the moment, so you could offer to pay 3 months upfront in cash -- and you'll sign whatever paperwork you need to, or agree to any contract, that gives the appearance that the actual rent on the place is $2000. Why not sweeten the deal and sign a stipulation saying that the landlord can charge you whatever she likes in a year and if you dispute it, the case gets automatically referred to a private arbitrator of the landlord's choosing?

That's what I would do if I were you and it was time to renegotiate a lease.