---
title: Why are NYC apartment prices not falling more?
author: Seth Ariel Green
date: '2020-05-21'
slug: why-are-nyc-apartment-prices-not-falling-more
---

### Observations:

* I have an apartment in New York. 
* I don't currently reside there and will move out when the lease ends (in about a month).
* Many people I know are in the same boat (and we know there's been something of an [exodus from the city](https://www.nytimes.com/interactive/2020/05/16/nyregion/nyc-coronavirus-moving-leaving.html)).

But prices don't seem to reflect that.

Browsing rentals on [StreetEasy](https://streeteasy.com/for-rent/nyc), prices seem to have fallen _slightly_ from where they were three months ago -- despite New York's being the epicenter of a pandemic, and therefore:

* the things that make it special (food, opera, whatever your cup of tea is) are basically in complete abeyance;
* the tradeoff you made to have a perfectly located apartment -- no direct sunlight but there are three ramen shops and two mezcal bars on your block and you spend most of your day out and about anyway -- suddenly sucks;
* everyday things like exiting and entering the building you share with elderly neighbors are much more fraught than they used to be;
* if [people](http://www.arnoldkling.com/blog/city-inhabitants-as-amenities/) are the [ultimate](http://www.arnoldkling.com/blog/the-coordination-problem-and-winner-cities/) [amenity](http://www.arnoldkling.com/blog/expensive-cities-and-labor-immobility/), a lot of them are gone.

So what gives?

My theory is that rent stabilization laws put a floor on how much much landlords are willing to let prices drop.

### Example:

* About a million apartments in New York are rent-stabilized if not rent controlled. 
* Let's say you (the landlord) have a rent-stabilized apartment that, in normal times, rents for \$2,000/month.
* The actual [market-clearing](https://en.wikipedia.org/wiki/Market_clearing) price for the apartment right now is \$1,500/month.
* You estimate that, over the next 10 years, you'll only be able to raise the rent an average of 5% per year (you can't get a perfect estimate of this because legislation is liable to change the rate).
* Two years from now, you expect the apartment to be worth $2,000/month again.

**What would you do?**

On the one hand, you could leave the apartment up on StreetEasy for \$2,000/month, even though you think that it will stay vacant for two years. On paper, you're going to lose \$48,000 in that time -- but remember, the true, market-clearing price for the aparmtent is $1,500/month, so you're actually losing \$36,000. Two years from now, you expect to rent it for \$2,000/month and increase the rent by 5% every year for the following 8 years.

On the other, you could reduce the price to $1,500 and then aim to raise the price by 5% every year for the next ten years. 

In the first scenario, your income for the 10 year period is:
```{r}
rent_calc <- function(base_amount, increase_rate, years) {
  rent <- (base_amount * 12)
  appreciation <- 1 + increase_rate
  for(i in 1:(years-1)){
    rent <- rent + (base_amount * appreciation^i * 12)
  }
  print(noquote(paste0("$", format(x = rent, big.mark = ',',
                                   trim = F, digits = 8))))
}

rent_calc(2000, 0.05, 8)
```

Alternatively:
```{r}
rent_calc(1500, 0.05, 10)
```

So you're a litlte better off leaving the apartment vacant for the next two years.

I've played it fast and loose with the details here -- rent-stabilization is complicated, applies only to certain buildings, vacant units aren't costless, etc. But if this story is basically right:

### A prediction and a suggestion

1) A lot of apartments will come on the market this year with perks like "3 months free," which effectively reduce rent by a lot, but, as far as I know, aren't typically covered by rent stabilization laws;

2) If you're renewing a lease, and you have a good relationship with your landlord, you could strike a sweet deal. For this hypothetical \$2,000 apartment, tell them that you're willing to pay \$1,000 a month, with the first 6 month upfront in cash, and you'll sign whatever deal they offer that puts the price of the apartment, on paper, at \$2,000 a month. If you really want to [tie your hands to the mast](https://spectacledavenger.blogspot.com/2010/08/jon-elster-ulysses-and-sirens.html), how about suggesting binding arbitration in the case of a dispute about any rent increase in a year's time?

That's what I would do if I were renegotiating a lease.